Contracts for unions and intersections are hard for Nickel

### 2021-06-04 paper.tex
Arnaud working more of his way through the introduction

On the other hand, unions are not a common feature of static type
systems, mostly because they are quite difficult to verify
statically. So unions are really only worth it in gradually typed
language where they formalise existing dynamically typed patterns. On
the other hand, surely, for dynamic tests, unions are really easy: it
is simply the boolean disjunctions of two tests.

@@

These are an inherently dynamic question, as they are all questions
about the evaluated configuration.

@@

We will use the syntax of Nickel to write most examples throughout this paper.
Syntactically, Nickel looks like a simple ML language with records. \nickel{let
id = value in exp} is used to bind a value to an identifier in an expression.
Functions are declared using the \nickel{fun arg1 ... argn => body} syntax. List
are written as \nickel{[x1, ..., xn]} and records as \nickel{\{field1 = value1,
..., fieldn = valuen\}}. The application of a contract \nickel{C} to the
expression \nickel{exp} is written \nickel{exp | C}\footnote{The current
    implementation requires a \nickel{#} character in front of custom contracts,
    whose application would rather be \nickel{exp | #C}, but it is irrelevant
and subject to change in the future.}. We will sometimes attach
contracts directly to an identifier instead of the corresponding bound
expression, that is: \nickel{let id | C = value in exp} is syntactic sugar for
\nickel{let id = (value | C) in exp}.

Some examples resort to constructions that are not currently available in Nickel
for illustration purpose: we will use \nickel{A @& B} (resp. \nickel{A @| B})
for the intersection (resp. union) of contracts (or types) \nickel{A} and
\nickel{B}. \nickel{@|} and \nickel{@&} bind more loosely than arrows, meaning
that \nickel{A -> B @& C -> D} is to be parsed as \nickel{(A -> B) @& (C -> D)}.

@@

% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall
% LocalWords:  Kubernetes DevOps natively

>>>>

On the other hand, unions are not a common feature of static type
systems, mostly because they are quite difficult to verify
statically. So unions are really only worth it in gradually typed
language where they formalise existing dynamically typed patterns. On
the other hand, surely, for dynamic tests, unions are really easy: it
is simply the Boolean disjunctions of two tests.

@@

These are inherently dynamic questions, as they are all questions
about the evaluated configuration. To this effect, Nickel lets us
annotate any expression with a dynamic schema check: \nickel{exp |
  C}. There is also syntactic sugar to annotate definitions:
\nickel{let id | C = value in exp} stands for \nickel{let id = (value
  | C) in exp}.

Let us pause for a moment and consider the following: it is Nickel's
ambition to be able to manipulate configurations like Nixpkgs. With
over 50\,000 packages, is one of the largest repository of software
packages in existence~\cite{repology}. Concretely, Nixpkgs is a
dictionary mapping packages to build recipes. That is, a massive,
over-50\,000-key-value-pair wide dictionary. It is absolutely out of the
question to evaluate the entirety of this dictionary every time one
needs to install 10 new packages: this would result in a painfully
slow experience.

To be able to support such large dictionaries, Nickel's dictionaries
are \emph{lazy}, that is, the values are only evaluated when
explicitly required. For instance, when writing
\hbox{\nickel{nixpkgs.hello},} only the \nickel{hello} package gets
evaluated.

But let's consider now writing something like \nickel{nixpkgs |
  packages}, to guarantee that all the packages conform to the desired
schema. If this were a simple Boolean test, it would have to evaluate
all 50\,000 package to check their validity, hence breaking the
laziness of dictionaries. Do we have to choose between laziness and
schema validation? Fortunately, we don't! Enter
\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic check
which can be partially delayed, yet errors can be reported
accurately. Contracts can respect laziness of dictionaries, and they
can be use to add schema validation to functions as well (in fact
functions were their original motivations).

The is no Boolean function which can check that a value has type

\rule{3cm}{3pt}

@@

@@

% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall MyPy
% LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
% LocalWords:  TypeScript
